\vspace{-3em} \textbf{ Mapping stellar surfaces\\ II: An interpretable Gaussian process model for light curves }

### 2021-03-30 paper2/ms.tex
reviewer comments; push dataset to zenodo

\defcitealias{PaperI}{Paper I}

@@

A major source of stellar variability in both light curves and radial velocity
datasets is the modulation induced by magnetically-driven surface features like starspots
rotating in and out of view. While GPs excel at marginalizing over stellar rotational
variability, they have been less useful when the goal is to make
inferences about the actual source of this variability, such as the
properties of starspots and the magnetic processes that generate them.
\textbf{While it is straightforward to derive
    posterior constraints on the hyperparameters of an effective GP model for
    observations of a star, it is not clear what those constraints actually
    tell us about the stellar surface.}
Specifically, in all but a few restricted cases, there is no \emph{first principles}
relationship between the descriptive parameters of a typical GP model
(see \S\ref{sec:gp-intro})
and the physical properties of the stellar surface that is being observed.
%
For instance, it may be tempting to interpret the GP amplitude hyperparameter
as some measure of the spot contrast or the total number of spots,
or the GP timescale hyperparameter as the
spot lifetime, but there are no guarantees these interpretations will
hold in general. After all, the choice of kernel is quite often \emph{ad hoc},
providing an \emph{effective}---as opposed to \emph{interpretable}---description
of the physics.
%
There are two important exceptions to this: asteroseismic studies, in which the
the GP hyperparameters can offer direct insight into
the behavior of complex pulsation modes and thus physical properties of the
stellar interior \citep[e.g.,][]{BrewerStello2009,ForemanMackey2017}; and
stellar rotation period studies, in which the period hyperparameter $P$ can usually
be associated with the rotation period of the star \citep[e.g.,][]{Angus2018}.%
\footnote{An exception to this is in the presence of strong differential
    rotation, in which case many periods may be present in the data, or
    when spots evolve coherently, which can also introduce weak periodicities
    in the light curve.}
%
For spot-induced variability, on the other hand,
GPs are usually used
when the variability itself is a nuisance parameter. For example, if the
goal is to constrain the properties of a transiting exoplanet or to
search for a planetary signal in a radial velocity dataset, a GP might be
used to remove (or, better yet, to marginalize over) the stellar variability
\citep[e.g.,][]{Haywood2014,Rajpaul2015,Luger2017b}.
In this case, the physics behind the variability is irrelevant, so an
effective model of this sort may be sufficient.

@@

        \pmb{\mu}(P, \mathbf{u}, \pmb{\theta}_\bullet)

@@

        \pmb{\Sigma}(P, \mathbf{u}, \pmb{\theta}_\bullet)

@@

We use Equation~(\ref{eq:log-like}) as our log likelihood
term, adding the log of the absolute value of Equation~(\ref{eq:J}) to
enforce a uniform prior on $\mu_\phi$ and $\sigma_\phi$. As we mentioned
above, the fact that $P$, $\mathbf{u}$, and $\sigma_f$ are shared among
all $M$ light curves means that $\pmb{\Sigma} + \mathbf{C}_m$ is the same for all
of them, greatly speeding up the likelihood evaluation since we need
only invert (or factorize) it a single time per sample. Our
covariance is the covariance of the \emph{normalized} process,
given by Equation~(\ref{eq:SigmaTilde}). Since we only consider light curves
with variability limited to a few percent or less, the approximation for
$\tilde{\pmb{\Sigma}}$ is always valid. Finally, we restrict our spherical
harmonic expansion to $l_\mathrm{max} = 15$ as a compromise between
resolution, computational speed, and numerical stability \citep[see][]{JOSSPaper}.

@@

Figure~\ref{fig:calibration_ld}---\ref{fig:calibration_ld_500_no_ld}
show results for the default run but with limb darkening. In all cases
we assume quadratic limb darkening with fiducial values $u_1 = 0.5$
and $u_2 = 0.25$ for all stars. From Figure~\ref{fig:calibration_ld}, in which we
assume we know the limb darkening coefficients exactly, it is clear that
the presence of limb darkening significantly degrades our ability to
infer both the radii and latitudes of the spots. Limb darkening has a complicated
effect on the mapping between surface features and disk-integrated flux, as it
reveals information about odd harmonics at
the expense of introducing strong degeneracies with the even harmonics
\citepalias{PaperI}. In practice, this leads to higher uncertainty in
the spot radii and latitudes relative to the same dataset without limb
darkening (Figure~\ref{fig:calibration_default_corner}). Fortunately,
this uncertainty can be dramatically reduced with more data, as evident
in Figure~\ref{fig:calibration_ld_1000}, which shows the results of the
same run but with $M=1{,}000$ light curves in the ensemble. The constraints
on $r$, $\mu_\phi$, and $\sigma_\phi$ are now much tighter and in
good agreement with the truth. Finally, Figure~\ref{fig:calibration_ld_500_no_ld}
shows the results of inference on limb-darkened light curves under the
(wrong) assumption that limb darkening is not present ($\mathbf{u} = \mathbf{0}$).
Neglecting the effect of limb darkening can lead to biases in the spot radius
and latitude parameters. While the model still favors mid-latitude spots
(at $\sim 45^\circ$ instead of $30^\circ$), the constraints are deceptively tight
and discrepant by many standard deviations. We discuss these points in
more detail in \S\ref{sec:caveats}.

@@

Because our expression for the GP covariance has an exact, closed-form
solution as a function of the spot parameters, it can be computed
efficiently: a typical likelihood evaluation on a dataset
consisting of $K \sim 1{,}000$ points takes about 20ms on a modern
laptop. Our algorithm is implemented in the open-source, user-friendly
\Python package \starryprocess, which is \textsf{pip}-installable, available on
\href{https://github.com/rodluger/starry_process}{GitHub}, and
described in \citet{JOSSPaper}.
The algorithm is implemented in a combination of \cpp and \Python,
linked using the \theano package. Because our GP covariance has an exact
representation, so too do its derivatives. We therefore implement
backpropagated derivatives with respect to all input parameters for
out-of-the-box usage with gradient-based inference and optimization
tools such as Hamiltionain Monte Carlo (HMC), autodifferentation
variational inference (ADVI), and gradient-based nested sampling.

@@

    \item \textbf{Our GP can be used in exoplanet transit modeling and

@@

\subsection{Basic expression}
%
\label{sec:starry:basic}
As we mentioned in \S\ref{sec:gp-gp}, the flux $\mathbf{f}$ is a purely
linear function of the spherical harmonic coefficient vector $\mathbf{y}$:
%
\begin{linenomath}\begin{align}
        \mathbf{f} = \mathbf{1} + \pmb{\mathcal{A}} \, \mathbf{y}
        \quad.
    \end{align}\end{linenomath}
%
Even though this is derived in detail in \citet{Luger2019}, it is useful to
expand on the computation of the design matrix $\pmb{\mathcal{A}}$ in more
detail here. Let $\mathbf{a}_k^\top$ denote the $k^{th}$ row of $\pmb{\mathcal{A}}$,
such that
%
\begin{linenomath}\begin{align}
        \label{eq:Arows}
        \pmb{\mathcal{A}}
         & =
        \setstackgap{L}{1.25\baselineskip}
        \fixTABwidth{T}
        \parenMatrixstack{
            \mathbf{a}_0^\top \\
            \mathbf{a}_1^\top \\
            \vdots            \\
            \mathbf{a}_{K-1}^\top
        }\quad.
    \end{align}\end{linenomath}
%
The row vector $\mathbf{a}_k^\top$ encodes how the spherical harmonic
coefficient vector projects onto the $k^\mathrm{th}$ cadence in the flux time series, and
may be computed from
%
\begin{linenomath}\begin{align}
        \label{eq:akT}
        \mathbf{a}_k^\top = \mathbf{r}^\top \,
        \mathbf{A_1} \,
        \mathbf{R}_{\hat{\mathbf{x}}}\left(-I\right) \,
        \mathbf{R}_{\hat{\mathbf{z}}}\left(\frac{2\pi}{P}t_k\right) \,
        \mathbf{R}_{\hat{\mathbf{x}}}\left(\frac{\pi}{2}\right)
        \quad.
    \end{align}\end{linenomath}
%
To understand the expression above, let us proceed from right to left,
starting with the spherical harmonic vector $\mathbbb{y}$, which we assume
describes the surface intensity of the star at time $t = 0$
in a frame where $\hat{\mathbf{x}}$
points to the right, $\hat{\mathbf{y}}$ points up, and $\hat{\mathbf{z}}$
points out of the page. The quantity $\mathbf{R}_{\hat{\mathbf{x}}}$
is a Wigner rotation matrix
(described in detail in \S\ref{sec:wigner}),
which in this case rotates the spherical harmonic representation
of the star by an angle $\nicefrac{\pi}{2}$ counter-clockwise about $\hat{\mathbf{x}}$
such that the north pole of the star points along $\hat{\mathbf{z}}$. In this
frame, we apply a second Wigner rotation matrix, $\mathbf{R}_{\hat{\mathbf{z}}}$,
to rotate the star about $\hat{\mathbf{z}}$ counter-clockwise (i.e., eastward) by an angle
$\nicefrac{2\pi t_k}{P}$, where $P$ is the rotation period and $t_k$ is the
time at cadence $t$.
Next, we rotate the star by a \emph{clockwise} angle of $I$ about $\hat{\mathbf{x}}$,
where $I$ is the stellar inclination ($I = 0$ corresponding to a pole-on view and
$I = \nicefrac{\pi}{2}$ corresponding to an edge-on view). With this last rotation,
we are now in the observer's frame.%
\footnote{In principle, one last rotation could be performed about $\hat{\mathbf{z}}$
    to orient the projected disk of the star on the plane of the sky; however, the disk-integrated
    flux is independent of the rotation angle along the plane of the sky
    (which we refer to as the \emph{obliquity}), so this step is unnecessary.}

>>>>

\defcitealias{PaperI}{Paper I}
\newcommand{\edited}[1]{\textbf{#1}}

@@

A major source of stellar variability in both light curves and radial velocity
datasets is the modulation induced by magnetically-driven surface features like starspots
rotating in and out of view. While GPs excel at marginalizing over stellar rotational
variability, they have been less useful when the goal is to make
inferences about the actual source of this variability, such as the
properties of starspots and the magnetic processes that generate them.
\edited{While it is straightforward to derive
    posterior constraints on the hyperparameters of an effective GP model for
    observations of a star, it is not clear what those constraints actually
    tell us about the stellar surface.}
Specifically, in all but a few restricted cases, there is no \emph{first principles}
relationship between the descriptive parameters of a typical GP model
(see \S\ref{sec:gp-intro})
and the physical properties of the stellar surface that is being observed.
%
For instance, it may be tempting to interpret the GP amplitude hyperparameter
as some measure of the spot contrast or the total number of spots,
or the GP timescale hyperparameter as the
spot lifetime, but there are no guarantees these interpretations will
hold in general. After all, the choice of kernel is quite often \emph{ad hoc},
providing an \emph{effective}---as opposed to \emph{interpretable}---description
of the physics.
%
There are two important exceptions to this: asteroseismic studies, in which the
the GP hyperparameters can offer direct insight into
the behavior of complex pulsation modes and thus physical properties of the
stellar interior \citep[e.g.,][]{BrewerStello2009,ForemanMackey2017}; and
stellar rotation period studies, in which the period hyperparameter $P$ can usually
be associated with the rotation period of the star \citep[e.g.,][]{Angus2018}.%
\footnote{An exception to this is in the presence of strong differential
    rotation, in which case many periods may be present in the data, or
    when spots evolve coherently, which can also introduce weak periodicities
    in the light curve.}
%
For spot-induced variability, on the other hand,
GPs are usually used
when the variability itself is a nuisance parameter. For example, if the
goal is to constrain the properties of a transiting exoplanet or to
search for a planetary signal in a radial velocity dataset, a GP might be
used to remove (or, better yet, to marginalize over) the stellar variability
\citep[e.g.,][]{Haywood2014,Rajpaul2015,Luger2017b}.
In this case, the physics behind the variability is irrelevant, so an
effective model of this sort may be sufficient.

@@

@@

@@

We use Equation~(\ref{eq:log-like}) as our log likelihood
term, adding the log of the absolute value of Equation~(\ref{eq:J}) to
enforce a uniform prior on $\mu_\phi$ and $\sigma_\phi$. As we mentioned
above, the fact that $P$, $\mathbf{u}$, and $\sigma_f$ are shared among
all $M$ light curves means that $\pmb{\Sigma} + \mathbf{C}_m$ is the same for all
of them, greatly speeding up the likelihood evaluation since we need
only invert (or factorize) it a single time per sample.%
%
\footnote{\edited{Re-factorization would incur a computational cost
        on the order of 10ms per target per likelihood evaluation for the
        datasets we consider here.}}
%
Our
covariance is the covariance of the \emph{normalized} process,
given by Equation~(\ref{eq:SigmaTilde}). Since we only consider light curves
with variability limited to a few percent or less, the approximation for
$\tilde{\pmb{\Sigma}}$ is always valid. Finally, we restrict our spherical
harmonic expansion to $l_\mathrm{max} = 15$ as a compromise between
resolution, computational speed, and numerical stability \citep[see][]{JOSSPaper}.

@@

Figure~\ref{fig:calibration_ld}---\ref{fig:calibration_ld_500_no_ld}
show results for the default run but with limb darkening. In all cases
we assume quadratic limb darkening with fiducial values $u_1 = 0.5$
and $u_2 = 0.25$ for all stars. From Figure~\ref{fig:calibration_ld}, in which we
assume we know the limb darkening coefficients exactly, it is clear that
the presence of limb darkening significantly degrades our ability to
infer both the radii and latitudes of the spots. Limb darkening has a complicated
effect on the mapping between surface features and disk-integrated flux, as it
reveals information about odd harmonics at
the expense of introducing strong degeneracies with the even harmonics
\citepalias{PaperI}. In practice, this leads to higher uncertainty in
the spot radii and latitudes relative to the same dataset without limb
darkening (Figure~\ref{fig:calibration_default_corner}). Fortunately,
this uncertainty can be dramatically reduced with more data, as evident
in Figure~\ref{fig:calibration_ld_1000}, which shows the results of the
same run but with $M=1{,}000$ light curves in the ensemble. The constraints
on $r$, $\mu_\phi$, and $\sigma_\phi$ are now much tighter and in
good agreement with the truth. \textbf{While there is evidence for slight bias
    in the inferred radius, this is likely due to the difference between the model
    used to generate the spots and that used for inference, as discussed in the
    previous section.}
Finally, Figure~\ref{fig:calibration_ld_500_no_ld}
shows the results of inference on limb-darkened light curves under the
(wrong) assumption that limb darkening is not present ($\mathbf{u} = \mathbf{0}$).
Neglecting the effect of limb darkening can lead to biases in the spot radius
and latitude parameters. While the model still favors mid-latitude spots
(at $\sim 45^\circ$ instead of $30^\circ$), the constraints are deceptively tight
and discrepant by many standard deviations. We discuss these points in
more detail in \S\ref{sec:caveats}.

@@

Because our expression for the GP covariance has an exact, closed-form
solution as a function of the spot parameters, it can be computed
efficiently: a typical likelihood evaluation on a dataset
consisting of $K \sim 1{,}000$ points takes \edited{between 30ms and 60ms} on a modern
laptop. Our algorithm is implemented in the open-source, user-friendly
\Python package \starryprocess, which is \textsf{pip}-installable, available on
\href{https://github.com/rodluger/starry_process}{GitHub}, and
described in \citet{JOSSPaper}.
The algorithm is implemented in a combination of \cpp and \Python,
linked using the \theano package. Because our GP covariance has an exact
representation, so too do its derivatives. We therefore implement
backpropagated derivatives with respect to all input parameters for
out-of-the-box usage with gradient-based inference and optimization
tools such as Hamiltionain Monte Carlo (HMC), autodifferentation
variational inference (ADVI), and gradient-based nested sampling.

@@

    \item \textbf{Our GP can be used in exoplanet transit modeling and extended to

@@

\subsection{Basic expression}
%
\label{sec:starry:basic}
As we mentioned in \S\ref{sec:gp-gp}, the flux $\mathbf{f}$ is a purely
linear function of the spherical harmonic coefficient vector $\mathbf{y}$:
%
\begin{linenomath}\begin{align}
        \label{eq:f1Ay}
        \mathbf{f} = \mathbf{1} + \pmb{\mathcal{A}} \, \mathbf{y}
        \quad.
    \end{align}\end{linenomath}
%
\edited{In this expression, we assume $\mathbf{y}$
    describes the surface intensity of the star at time $t = 0$
    in a frame where $\hat{\mathbf{x}}$
    points to the right, $\hat{\mathbf{y}}$ points up, and $\hat{\mathbf{z}}$
    points out of the page.}
%
Even though this is derived in detail in \citet{Luger2019}, it is useful to
expand on the computation of the design matrix $\pmb{\mathcal{A}}$, \edited{which
    transforms from spherical harmonics to flux}.
Let $\mathbf{a}_k^\top$ denote the $k^{th}$ row of $\pmb{\mathcal{A}}$,
such that
%
\begin{linenomath}\begin{align}
        \label{eq:Arows}
        \pmb{\mathcal{A}}
         & =
        \setstackgap{L}{1.25\baselineskip}
        \fixTABwidth{T}
        \parenMatrixstack{
            \mathbf{a}_0^\top \\
            \mathbf{a}_1^\top \\
            \vdots            \\
            \mathbf{a}_{K-1}^\top
        }\quad.
    \end{align}\end{linenomath}
%
The row vector $\mathbf{a}_k^\top$ encodes how the spherical harmonic
coefficient vector projects onto the $k^\mathrm{th}$ cadence in the flux time series, and
may be computed from
%
\begin{linenomath}\begin{align}
        \label{eq:akT}
        \mathbf{a}_k^\top = \mathbf{r}^\top \,
        \mathbf{A_1} \,
        \mathbf{R}_{\hat{\mathbf{x}}}\left(-I\right) \,
        \mathbf{R}_{\hat{\mathbf{z}}}\left(\frac{2\pi}{P}t_k\right) \,
        \mathbf{R}_{\hat{\mathbf{x}}}\left(\frac{\pi}{2}\right)
        \quad.
    \end{align}\end{linenomath}
%
To understand the expression above, let us \edited{consider how each of its terms operate on
    the spherical harmonic representation $\mathbf{y}$} from right to left.
The quantity $\mathbf{R}_{\hat{\mathbf{x}}}$
is a Wigner rotation matrix
(described in detail in \S\ref{sec:wigner}),
which in this case rotates the spherical harmonic representation
of the star by an angle $\nicefrac{\pi}{2}$ counter-clockwise about $\hat{\mathbf{x}}$
such that the north pole of the star points along $\hat{\mathbf{z}}$. In this
frame, we apply a second Wigner rotation matrix, $\mathbf{R}_{\hat{\mathbf{z}}}$,
to rotate the star about $\hat{\mathbf{z}}$ counter-clockwise (i.e., eastward) by an angle
$\nicefrac{2\pi t_k}{P}$, where $P$ is the rotation period and $t_k$ is the
time at cadence $t$.
Next, we rotate the star by a \emph{clockwise} angle of $I$ about $\hat{\mathbf{x}}$,
where $I$ is the stellar inclination ($I = 0$ corresponding to a pole-on view and
$I = \nicefrac{\pi}{2}$ corresponding to an edge-on view). With this last rotation,
we are now in the observer's frame.%
\footnote{In principle, one last rotation could be performed about $\hat{\mathbf{z}}$
    to orient the projected disk of the star on the plane of the sky; however, the disk-integrated
    flux is independent of the rotation angle along the plane of the sky
    (which we refer to as the \emph{obliquity}), so this step is unnecessary.}
